One-arm spirals in locally isothermal, radially structured self-gravitating discs

### 2014-11-12 paper/discussion.tex
minor updates

\subsection{The confined spiral as an external potential}
One possible effect of self-gravity is to allow the one-arm spiral 
to act as an external potential for the wider disc. This effect is
analogous to disc-satellite interaction 
\citep{goldreich79}, and may lead to instability 
if the angular momentum associated with the spiral has the opposite 
sign to that of the density waves it induces in the exterior disc
\citep{lin11b}. To estimate the magnitude of this effect, we port basic results from
disc-planet theory \citep[see, e.g.][and references
therein]{papaloizou07}.

Let us treat the one-arm spiral disturbance located in the inner disc as an  
external potential of the form $\Phi_\mathrm{ext}(R)\cos{\left(\phi -
    \Omega_pt\right)}$. Let $M_\mathrm{ring}$ denote the
mass contained within $R\in[R_1,R_2]$ of the disc, where the spiral is
confined in the numerical simulations. We take 
\begin{align} 
  \Phi_\mathrm{ext}
  =-\frac{GM_\mathrm{ring}}{\bar{R}}b^{1}_{1/2}(\beta),   
\end{align}
where $b_{n}^m(\beta)$ is the Laplace coefficient, $\beta = R/\bar{R}$ 
\citep{goldreich79} and $\bar{R} = (R_1+R_2)/2$ is the
approximate location of the spiral.

%and here we are interested in the
%potential it produces at $R>\bar{R}$  
%This potential is associated with a mass
%$M_\mathrm{ring}$ contained in   is the mass in $R\in[R_1,R_2]$.

%For an order-of-magnitude exercise, we simply take 
%\begin{align}
 % \Phi_\mathrm{ext} = -\frac{GM_\mathrm{ring}}{|R - \bar{R}|}, 
%\end{align}
%for the potential associated with a disturbance located at
%$R=\bar{R}$, where $M_\mathrm{ring}$

We expect the external potential to exert a torque at the Lindblad  
and co-rotation resonances. At the outer Lindblad resonance (OLR), this
torque is 
\begin{align}
  \Gamma_L =
  \frac{\pi^2\Sigma_L}{3\Omega_L\Omega_p}
\left[\left.R_L\frac{d\Phi_\mathrm{ext}}{dR}\right|_L + 2\left(1 - \frac{\Omega_p}{\Omega_L}\right)\Phi_\mathrm{ext}\right]^2,
\end{align}   
where a Keperian disc has been assumed and subscript $L$ denotes
evaluation at the OLR, $R=R_L$. (The inner Lindblad resonance does not
exist for the pattern speeds observed in our simulations.)

If we also associate its angular momentum with 
$J_\mathrm{ext}  = M_\mathrm{ring}\bar{R}^2\Omega_p$, we can obtain a
rate $\gamma_L=\Gamma_L/J_\mathrm{ext}$. Then 
\begin{align}
  \frac{\gamma_L}{\Omega_p} = &\frac{\pi
    h}{3Q_L}\left(\frac{M_p}{M_*}\right)\left(\frac{R_L}{\bar{R}}\right)\left(\frac{R_c}{\bar{R}}\right)^3\left(\frac{R_L}{R_c}\right)^{-3/2}\notag\\
  &\times \left\{\frac{R_L}{\bar{R}}\left.\frac{db_{1/2}^1}{d\beta}\right|_L + 2\left[1 -
      \left(\frac{R_c}{R_L}\right)^{-3/2}\right]b_{1/2}^1(\beta_L)\right\}^2.
\end{align}

%\begin{align}
 % \frac{\gamma_L}{\Omega_p} \sim &\frac{\pi h}{3
  %  Q_L}\left(\frac{M_\mathrm{ring}}{M_*}\right)\left(\frac{R_L}{\bar{R}}\right)^{-1/2}\left(\frac{R_c}{\bar{R}}\right)^{5/2}\notag\\
 % &\times \left\{ \frac{R_L}{R_L - \bar{R}} - 2\left[1 -
  %    \left(\frac{R_c}{R_L}\right)^{-3/2}\right]\right\}^2\frac{R_c^2}{\left(R_L
   % - \bar{R}\right)^2}. 
%\end{align}

Inserting $h=0.05$, $Q_L=10$, $M_\mathrm{ring} = 0.05M_*$,
$R_L=7.2R_0$, $R_c=4.4R_0$ and $\bar{R}=1.5R_0$ from our fiducial
FARGO simulation, we get

\begin{align}
  \gamma_L \sim 0.0005\Omega_p. 
\end{align}

%\begin{align}
 % \gamma_L \sim 0.012\Omega_p. 
%\end{align}

%This represents an upper limit on the effect of the OLR on the spiral
%disturbance, and is much smaller than the growth rate measured in the
%simulation ($\gamma\sim0.1\Omega_p$). We conclude that for our models,
%the OLR has negligible effect on the one-arm spiral.

For the co-rotation torque, we use the result
\begin{align}
  \Gamma_C = \left.
    \pi^2\Phi_\mathrm{ext}^2\left(\frac{d\Omega}{dR}\right)^{-1}\frac{d}{dR}\left(\frac{2\Sigma}{\Omega}\right)\right|_{C}
\end{align}
for a Keplerian disc, where subscript $C$ denotes evaluation at
co-rotation radius $R=R_c$. For a power-law surface density profile
$\Sigma\propto R^{-s}$ we have
\begin{align}
\frac{\gamma_c}{\Omega_p} = \frac{4}{3}\frac{\pi h}{Q_c}
\left(\frac{M_\mathrm{ring}}{M_*}\right)\left(\frac{R_c}{\bar{R}}\right)^4\left(s-\frac{3}{2}\right) \left[b_{1/2}^1(\beta_c)\right]^2 
\end{align}

%\begin{align}
%  \frac{\Gamma_C}{J_\mathrm{ext}\Omega_p} = \frac{4}{3}\frac{\pi h}{Q_C} \left(s -
 %   \frac{3}{2}\right)\left(\frac{M_\mathrm{ring}}{M_*}\right)\frac{R_c^4}{\bar{R}^2\left(R_c
 %     - \bar{R}\right)^2}.   
%\end{align}

Using the above parameter values with $s=2$ and $Q_c=10$, we obtain a
rate
\begin{align}
  \gamma_c\sim 0.0006\Omega_p,
\end{align}

%\begin{align}
%  \gamma_C\sim 0.010\Omega_p,
%\end{align}
which is again small compared to the measured growth rates.

>>>>

\subsection{The confined spiral as an external potential}
One possible effect of self-gravity is to allow the one-arm spiral  
to act as an external potential for the wider disc. This is
analogous to disc-satellite interaction 
\citep{goldreich79}, and may lead to instability 
if the angular momentum associated with the spiral has the opposite 
sign to that of the density waves it induces in the exterior disc
\citep{lin11b}. Here, we estimate the magnitude of this effect using
basic results from disc-planet theory \citep[see, e.g.][and references 
therein]{papaloizou07}.

Let us treat the one-arm spiral confined in $R\in[R_1,R_2]$ as an  
external potential of the form $\Phi_\mathrm{ext}(R)\cos{\left(\phi -
    \Omega_pt\right)}$. We take 
\begin{align} 
  \Phi_\mathrm{ext}
  =-\frac{GM_\mathrm{ring}}{\overline{R}}b^{1}_{1/2}(\beta),   
\end{align}
where $M_\mathrm{ring}$ is the disc mass contained within
$R\in[R_1,R_2]$, $\overline{R} = (R_1+R_2)/2$ is the approximate radial
location of the spiral, $b_{n}^m(\beta)$ is the Laplace coefficient
and $\beta = R/\overline{R}$. This form of
$\Phi_\mathrm{ext}$ is the $m=1$ component of the gravitational
potential of an external satellite on a circular orbit
\citep{goldreich79}.

%and here we are interested in the
%potential it produces at $R>\overline{R}$  
%This potential is associated with a mass
%$M_\mathrm{ring}$ contained in   is the mass in $R\in[R_1,R_2]$.

%For an order-of-magnitude exercise, we simply take 
%\begin{align}
 % \Phi_\mathrm{ext} = -\frac{GM_\mathrm{ring}}{|R - \overline{R}|}, 
%\end{align}
%for the potential associated with a disturbance located at
%$R=\overline{R}$, where $M_\mathrm{ring}$

We expect the external potential to exert a torque on the disc at the
Lindblad and co-rotation resonances. At the outer Lindblad resonance
(OLR), this torque is 
\begin{align}
  \Gamma_L =
  \frac{\pi^2\Sigma_L}{3\Omega_L\Omega_p}
  \left[\left.R_L\frac{d\Phi_\mathrm{ext}}{dR}\right|_L + 2\left(1 - \frac{\Omega_p}{\Omega_L}\right)\Phi_\mathrm{ext}\right]^2,
\end{align}   
where a Keperian disc has been assumed and subscript $L$ denotes
evaluation at the OLR, $R=R_L$. (The inner Lindblad resonance does not
exist for the pattern speeds observed in our simulations.)

If we associate the external potential with angular momentum 
$J_\mathrm{ext}  = M_\mathrm{ring}\overline{R}^2\Omega_p$, we can obtain a
rate $\gamma_L=\Gamma_L/J_\mathrm{ext}$. Then 
\begin{align}
  \frac{\gamma_L}{\Omega_p} = &\frac{\pi
    h}{3Q_L}\left(\frac{M_p}{M_*}\right)\left(\frac{R_L}{\overline{R}}\right)\left(\frac{R_c}{\overline{R}}\right)^3\left(\frac{R_L}{R_c}\right)^{-3/2}\notag\\
  &\times \left\{\frac{R_L}{\overline{R}}\left.\frac{db_{1/2}^1}{d\beta}\right|_L + 2\left[1 -
      \left(\frac{R_c}{R_L}\right)^{-3/2}\right]b_{1/2}^1(\beta_L)\right\}^2.
\end{align}
%\begin{align}
 % \frac{\gamma_L}{\Omega_p} \sim &\frac{\pi h}{3
  %  Q_L}\left(\frac{M_\mathrm{ring}}{M_*}\right)\left(\frac{R_L}{\overline{R}}\right)^{-1/2}\left(\frac{R_c}{\overline{R}}\right)^{5/2}\notag\\
 % &\times \left\{ \frac{R_L}{R_L - \overline{R}} - 2\left[1 -
  %    \left(\frac{R_c}{R_L}\right)^{-3/2}\right]\right\}^2\frac{R_c^2}{\left(R_L
   % - \overline{R}\right)^2}. 
%\end{align}
Inserting $h=0.05$, $Q_L=10$, $M_\mathrm{ring} = 0.05M_*$,
$R_L=7.2R_0$, $R_c=4.4R_0$ and $\overline{R}=1.5R_0$ from our fiducial
FARGO simulation, we get
\begin{align}
  \gamma_L \sim 5\times10^{-4}\Omega_p. 
\end{align}
%\begin{align}
 % \gamma_L \sim 0.012\Omega_p. 
%\end{align}
%This represents an upper limit on the effect of the OLR on the spiral
%disturbance, and is much smaller than the growth rate measured in the
%simulation ($\gamma\sim0.1\Omega_p$). We conclude that for our models,
%the OLR has negligible effect on the one-arm spiral.  
For the co-rotation torque, we use the result
\begin{align}
  \Gamma_c = \left.
    \pi^2\Phi_\mathrm{ext}^2\left(\frac{d\Omega}{dR}\right)^{-1}\frac{d}{dR}\left(\frac{2\Sigma}{\Omega}\right)\right|_{c}   
\end{align}
for a Keplerian disc, where subscript $c$ denotes evaluation at
co-rotation radius $R=R_c$. For a power-law surface density profile
$\Sigma\propto R^{-s}$ we have
\begin{align}
\frac{\gamma_c}{\Omega_p} = \frac{4}{3}\frac{\pi h}{Q_c}
\left(\frac{M_\mathrm{ring}}{M_*}\right)\left(\frac{R_c}{\overline{R}}\right)^4\left(s-\frac{3}{2}\right)
\left[b_{1/2}^1(\beta_c)\right]^2   
\end{align}
%\begin{align}
%  \frac{\Gamma_C}{J_\mathrm{ext}\Omega_p} = \frac{4}{3}\frac{\pi h}{Q_C} \left(s -
 %   \frac{3}{2}\right)\left(\frac{M_\mathrm{ring}}{M_*}\right)\frac{R_c^4}{\overline{R}^2\left(R_c
 %     - \overline{R}\right)^2}.   
%\end{align}
Using the above parameter values with $s=2$ and $Q_c=10$, we obtain a
rate
\begin{align}
  \gamma_c\sim 6\times 10^{-4}\Omega_p. 
\end{align}

These estimates are much smaller than that due to
the imposed temperature gradient as measured in the simulations
$\gamma\sim 0.1\Omega_p$). We conclude that for our disc models, the
Lindblad and co-rotation resonances have negligible effect on the
growth of the spiral arm in the inner disc. We confirmed this with
additional FARGO simulations with a reduced radial domain size,
setting $R_\mathrm{max}=3R_0$, which still developed the one-arm
spiral.

% both are positive -> destabilizing in principle

%\begin{align}
%  \gamma_C\sim 0.010\Omega_p,
%\end{align}
%which is again small compared to the measured growth rates.
